---
title: "Getting started with bayesqm"
author: "Raymond Dacosta Azadda"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with bayesqm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 110
)
```

This vignette walks through one Q study end to end in **bayesqm**:
import, fit, diagnose, interpret, select K, and export. The structure
mirrors the classical `qmethod` workflow so readers can map every
step onto its Bayesian analogue.

On your own machine, you would fit a real model with `fit_bayesian()`
via a Stan backend. For this document we work from a synthetic fit
produced by `demo_fit()` so every plot and summary below renders
without requiring Stan.

```{r, message = FALSE, warning = FALSE}
library(bayesqm)
```

## The Q-sort data

A Q study asks each participant to rank-order a set of statements
into a forced distribution. `bayesqm` represents that as a `J × N`
numeric matrix (statements as rows, participants as columns) plus a
vector of forced-distribution counts. For a real study, `read_qsort()`
auto-detects the file format:

```{r, eval = FALSE}
qdata <- read_qsort("mystudy.csv")   # CSV / Excel / HTMLQ / FlashQ
qdata <- read_qsort("mystudy.DAT")   # PQMethod
qdata <- read_qsort("mystudy.zip")   # KADE project
```

`qmethod` users can use the dotted aliases instead:
`import.pqmethod()`, `import.htmlq()`, `import.kenq()`,
`import.easyhtmlq()`.

For this vignette we simulate a dataset with `generate_data()` and
wrap it with `qsort_data()`:

```{r}
set.seed(1)
sim   <- generate_data(N = 20, J = 22, K = 2, seed = 1)
qdata <- qsort_data(sim$Y, distribution = sim$distribution,
                    source = "simulated")
qdata
```

## Fitting the model

On real data you would call `fit_bayesian()`:

```{r, eval = FALSE}
fit <- fit_bayesian(qdata, K = 2, chains = 4, iter = 2000,
                    warmup = 1000, seed = 1)
```

The function samples the posterior of a low-rank factor model with a
Student-t likelihood (by default) and a hierarchical normal prior on
the loadings, then resolves rotational ambiguity with MatchAlign.
For this vignette, we use the synthetic fit helper:

```{r}
fit <- demo_fit(N = 20, J = 22, K = 2, seed = 1)
fit
```

`summary(fit)` adds factor characteristics, the PSIS-LOO ELPD (when
available), distinguishing-statement counts, and the MatchAlign
diagnostic.

## Diagnostics

Convergence statistics live in `fit$diagnostics`:

```{r}
fit$diagnostics
```

`plot_tucker()` visualises the per-draw Tucker's phi between each
aligned factor column and the MatchAlign pivot. Values near 1 indicate
a stable alignment; bimodality signals residual label-switching.

```{r tucker, fig.cap = "MatchAlign alignment quality by factor."}
plot_tucker(fit)
```

## Factor loadings

Each participant has a posterior distribution over their loading on
every factor. `plot_loading_posterior()` draws the loading forest,
with nested 50 percent and 95 percent credible-interval whiskers and
the classical Brown cut-off as a faint reference. Filled points are
participants with `P(factor k dominant) > 0.5`.

```{r loadings, fig.cap = "Loadings with nested 50% and 95% credible intervals."}
plot_loading_posterior(fit)
```

The same information as a tidy data frame:

```{r}
head(compute_loadings(fit$Lambda_draws, prob = 0.95))
```

## Factor z-scores

`plot()` gives a cross-panel z-score dotchart; rows share an order
(by factor 1's posterior mean, configurable via `sort_by`) so the
reader can scan horizontally to compare factors.

```{r zscores, fig.cap = "Posterior z-scores across factors with 50 / 95 CrIs."}
plot(fit)
```

For a single statement across factors, `plot_zscore_posterior()` gives
the drill-down:

```{r zscore-one, fig.cap = "Posterior z-score for a single statement across factors."}
plot_zscore_posterior(fit, statement = 1)
```

## Choosing K

`run_bayes()` fits the model for each K in a range and applies the
**peak-plus-Sivula** protocol: the ELPD peak is the automated choice;
the Sivula (2025) parsimony rule is reported alongside.

```{r, eval = FALSE}
run <- run_bayes(qdata, K_max = 4, seed = 1,
                 chains = 2, iter = 1500, warmup = 800)
run
plot_elpd(run)
```

The `run$case` field labels this relationship. When both methods pick
the same K, the case is `agree` and the data supports that K
unambiguously. When the ELPD peak exceeds the Sivula choice, the case
is `gap`: the richer model fits a bit better on LOO but not
decisively, and the choice between them should turn on substantive
evidence. When Sivula picks a larger K than the peak (`reversed`),
something has usually gone wrong in the ELPD comparison itself; this
case is rare in practice and should not be read as a substantive
finding.

## Distinguishing, consensus, and membership

`bayesqm` replaces the classical z-score test with a direct posterior
probability. For every statement and every factor pair,
`P(|f_jk − f_jl| > δ)` answers *how confidently do the factors
separate this statement?* The default `δ = 1` is on the standardised
factor-score scale.

```{r dist-cons, fig.cap = "Distinguishing-statement probability. Rows sorted so the most discriminating statements are at the top."}
plot_dist_cons(fit, delta = 1.0)
```

Participant-level membership is probabilistic too.
`classify_membership()` turns posterior dominance into a
`Strong / Moderate / Weak` tier:

```{r}
head(classify_membership(fit$Lambda_draws))
```

```{r membership, fig.cap = "Dominant-factor posterior probability with tier strip."}
plot_membership(fit)
```

## Posterior predictive check

`plot_ppc()` shows the posterior distribution of the RMSE between
`cor(Y_rep)` and `cor(Y_obs)`. A well-specified model puts most mass
at small RMSE.

```{r ppc, fig.cap = "PPC on the correlation matrix."}
plot_ppc(fit)
```

## Hyperparameters

```{r hyper, fig.cap = "Posterior densities of nu, sigma, and tau."}
plot_hyper(fit)
```

## Reporting and exporting

`save_bayesqm_plot()` writes any plot to PDF / SVG / PNG / TIFF /
JPEG at journal-appropriate dimensions:

```{r, eval = FALSE}
save_bayesqm_plot("fig_loadings.pdf", plot_loading_posterior(fit))
save_bayesqm_plot("fig_elpd.pdf",      plot_elpd(run),
                  width = 3.5, height = 3)
```

`caption_bayesqm()` returns a ready-to-paste figure caption with
model configuration, sample sizes, interval probability, and
convergence diagnostics:

```{r}
caption_bayesqm(fit)
```

Standard R accessors work on the fit (`coef`, `fitted`, `residuals`,
`sigma`, `family`, `as.matrix`), and `as.matrix(fit)` returns a
Stan-style draws matrix that `posterior` and `bayesplot` consume
natively.

## Renaming factors and theming

Substantive factor labels replace `f1..fK` in every slot:

```{r}
fit2 <- rename_factors(fit, c("tradition", "innovation"))
plot(fit2)
```

Every plot reads its palette through `bayesqm_colors()`; switch the
scheme once and every subsequent plot follows:

```{r, eval = FALSE}
bayesqm_set_colors("teal")   # "blue" (default), "red", "purple", "grey"
plot(fit)
bayesqm_set_colors("blue")
```

## Where next

- `?fit_bayesian` — every prior and sampler option
- `?run_bayes` — peak-plus-Sivula thresholds
- `?bayesqm-membership` — the full set of membership summaries
- `ggplot2::autoplot()` — ggplot2 / ggdist versions of every view
  above, when `ggplot2` and `ggdist` are installed
